Light inextensible strings under tension in the Schwarzschild geometry

### 2019-08-28 string_physics_teacher.tex
remove the tangential-to-the-event-horizon stuff

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} but it is a straightforward second order ODE,
amenable to numerical solution.  There are two qualitatively different
types of solutions: those with a tangential point ($r'=0,r>2M$), and
those without.  Consider first solutions with such a point: there
exists a pair~$\left(r,\phi\right), r>2M$ at which~$r'(\phi)=0$.  This
point must be unique, as~$r''(\phi)>0$, and be the closest approach of
the string to the event horizon.

Further, because the system is invariant
under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
at this point.  Considering~$r'(\phi)=0$ shows that a tangential
string at a point with~$r_0>2M$ obeys, to first order, $r''=r-2M$,
while tangential null geodesics follow $r''=r-3M$.  Thus a taut string
would {\em look} curved to a local observer, having a (visual) radius
of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from
the fact that appearance is driven by local $ds$, not Schwarzschild
$r$).  At the photon sphere $r_0=3M$, such a string would appear to
have a radius of~$\sqrt{27}M$.  For a solar mass black hole this would
be about $15\unit{km}$, a readily understandable result.

If, conversely, $r'(\phi)\neq 0$ whenever~$r>2M$, then the string has
only one free end; the other is tangential to the event horizon (in
the sense that ${\displaystyle \lim_{r\longrightarrow 2M^+}r'=0}$),
except in the degenerate case of constant~$\phi$ when the string is
radial for~$r>2M$.

@@

There are other ways to parametrize completely exterior strings;
Figure \ref{strings_r_equals_2}, for example, shows strings passing
through~$(4M,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
comparison.  The cases where the string attains $r=2M$ are tangent to
the event horizon (or, exceptionally, purely radial).
Figure~\ref{fixed_EH_intersection} shows strings tangential to the
event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
shows tangential strings passing through $(4M,0)$.

@@

\begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"
\centering
\includegraphics[width=\linewidth]{one_free_end_fixed_EH_intersection.pdf}
\caption{Strings tangential to event horizon at $(1,0)$}
\label{fixed_EH_intersection}
\end{figure}

\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
\centering
\includegraphics[width=\linewidth]{one_free_end_r_equals_2.pdf}
\caption{Strings tangential to the event horizon passing through $(4M,0)$}
\label{one_free_end_r_equals_2}
\end{figure}

>>>>

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} but it is a straightforward second order ODE,
amenable to numerical solution.  Here we consider only strings
exterior to the event horizon, $r>2M$, and for these we have a
well-defined point of closest approach to the event horizon.  Further,
because the system is invariant under~$\phi\longrightarrow-\phi$, the
string must be symmetrical about at this point.
Considering~$r'(\phi)=0$ shows that a tangential string at a point
with~$r_0>2M$ obeys, to first order, $r''=r-2M$, while tangential null
geodesics follow $r''=r-3M$.  Thus a taut string would {\em look}
curved to a local observer, having a (visual) radius of curvature of
$r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the fact that
appearance is driven by local $ds$, not Schwarzschild $r$).  At the
photon sphere $r_0=3M$, such a string would appear to have a radius
of~$\sqrt{27}M$.  For a solar mass black hole this would be about
$15\unit{km}$, a readily understandable result.

@@

There are other ways to parametrize completely exterior strings;
Figure \ref{strings_r_equals_2}, for example, shows strings passing
through~$(4M,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
comparison.

@@
